This work establishes the EnKF as a dynamic and mechanistically grounded framework for knowledge transfer in upstream bioprocess modelling, allowing a system-specific mechanistic model to be progressively adapted to new conditions without reparametrisation. A central contribution lies in the dynamic evolution of parameter ensembles, which provide a temporally resolved view of how the model adapts to the new process. Unlike traditional approaches that treat model parameters as static, the EnKF framework tracks how parameter estimates evolve in response to real-time measurements. This time-resolved sensitivity analysis reveals which parameters are the most influential under new conditions, and when their impact on model behaviour becomes significant. Importantly, this is achieved without introducing artificial variability through parameter covariance inflation, thereby ensuring that changes in ensemble spread arise from real process data only.

Preserving mechanistic interpretability is particularly valuable for knowledge transfer, as it allows models to explain system behaviour beyond predictive performance. While hybrid and black-box approaches often prioritise generalisability, recent studies demonstrate that full mechanistic transparency can still be retained across diverse transfer contexts. Richelle \textit{et al.} demonstrated transfer from fed-batch to perfusion by simplifying model structure and introducing a ‘catch-all’ by-product variable to accommodate changes in operating mode \cite{Richelle2022Model-basedPerfusion}. Arndt \textit{et al.}  developed a scale-up workflow that quantified and compared model parameter distributions across bioreactor volumes, revealing when transfer is valid and when scale-dependent reparametrisation is necessary \cite{Arndt2021Model-basedSystems}. More recently, Lu \textit{et al.} incorporated time-series transcriptomics into a multi-cell-line learning (MCL) framework providing enzyme-level metabolic insight across CHO clones, though requiring high-performance computing and rich omics datasets \cite{Lu2024Multi-cell-lineModels}. This work complements these efforts by introducing time-resolved parameter trajectories, with the EnKF treating biologically meaningful kinetic parameters as dynamic quantities that evolve in response to process measurements. Compared with MCL, the presented method offers a computationally efficient alternative using only routine extracellular measurements, particularly suited for early process development where omics data may not be available.

{\color{blue} The ensemble-based formulation of the EnKF is inherently robust to the large model-measurement discrepancies that knowledge transfer entails. Conventional estimators such as the EKF and UKF were developed for settings in which the model is already reasonably accurate and corrections remain small. In knowledge transfer, major model-measurement discrepancies produce large corrections that can drive the explicitly propagated covariance matrix to become ill-conditioned or indefinite, making these estimators numerically unstable. The EnKF is substantially less susceptible to these failure modes as it represents uncertainty through the ensemble itself rather than an explicitly propagated covariance matrix. The state and parameter covariance is recomputed at each update directly from the ensemble, producing a sample covariance that is positive semi-definite by construction, with all variances guaranteed non-negative and no Jacobian linearisation required.}

While this method offers notable advantages in flexibility and interpretability, it relies on two key assumptions. First, the method is designed to capture process-specific differences through parameter adaptation, without changing the underlying model structure. {\color{blue} The six datasets validated here span two CHO cell lines under glutamine synthetase selection, Cell Line A (CHO-T127) producing an IgG1 and Cell Line B (CHO-GS46) producing an IgG4 monoclonal antibody, alongside two bioreactor scales, a temperature downshift, and three feeding strategies. The EnKF framework itself is not restricted to a specific system and can operate with any underlying mechanistic model. The critical requirement is that the model structure adequately represents the key biological phenomena of the target system. The EnKF framework can tolerate some degree of structural imperfection through parameter adaptation, as illustrated by the temperature downshift case where the absence of explicit temperature-dependent kinetics in the model is compensated by shifts in growth and productivity parameters that remain phenomenologically interpretable. However, when the target system involves phenomena that are structurally absent from the model, the method may still produce state estimates, but the unmodelled dynamics will be distributed across existing parameters in ways that cannot be meaningfully interpreted. Relevant examples include transferring between non-GS and GS-CHO systems without restructuring the glutamine and ammonia metabolism equations, scaling to pilot or manufacturing scale without incorporating mass transfer limitations, or applying the framework to a product modality with substantially different synthesis kinetics. In such cases, the model structure should be updated to reflect the biology of the new system before the framework is applied.} Second, the evolution of parameters is inherently conditioned on the initial ensemble drawn from the base system. In highly nonlinear systems, multiple parameter combinations may yield similar outputs, and the estimated trajectory represents one of the plausible solutions. These assumptions do not undermine the utility of the method but are essential to consider when interpreting biological meaning. Provided the model structure remains appropriate, the dynamic adaptation of parameters still offers valuable insight into system behaviour and sensitivity, especially when full reparametrisation is not feasible.

As the first demonstration EnKF dual state and parameter estimation in bioprocessing literatures, this work shows how dynamic model adaptation based on routine extracellular measurements can reveal system-specific behaviour without requiring full reparametrisation. The resulting dynamic parameter trajectories not only increase the flexibility of transferring mechanistic models across systems, but also provide biologically meaningful insight into process adaptation that extends beyond what is observable from concentration profiles alone. Beyond its methodological contributions, this framework offers a practical and interpretable route for reusing mechanistic models in new bioprocess settings where data are limited and timelines are constrained. From a regulatory perspective, the preservation of mechanistic structure and the transparent evolution of model parameters can support scientific justification for process comparability and risk assessment during tech transfer. In this context, the EnKF establishes itself as a robust and accessible tool for supporting knowledge-driven decision-making across development stages in upstream bioprocessing.