\label{SI:gsa_comparison}
{\color{blue}
To assess whether the EnKF identifies a consistent set of influential parameters without prior sensitivity information, the ten most-converging EnKF parameters (ranked by ensemble spread reduction across all six datasets; see Table~\ref{tab:spread_reduction}) are compared with the ten parameters identified by global sensitivity analysis (GSA) and subsequently re-estimated by Kotidis \textit{et al.}~\cite{Kotidis2019UseCultures} for Cell Line B (Table~\ref{tab:param_overlap}).

Seven of the ten parameters overlap (70\%), despite the EnKF receiving no sensitivity guidance. Both methods flag the core growth ($\mu_{\text{max}}$), lactate switching ($Lac_{\text{max1}}$, $Lac_{\text{max2}}$), amino acid yield ($Y_{X,\text{Asn}}$, $Y_{X,\text{Glu}}$, $Y_{X,\text{Amm}}$), and glucose maintenance ($m_{\text{Glc}}$) parameters as requiring adaptation for cross-cell-line transfer. Each method additionally identifies one mAb-production parameter, though a different one: $m_{\text{mAb}}$ (maintenance-associated) via GSA and $Y_{\text{mAb},\mu}$ (growth-associated) via EnKF. The three EnKF-only parameters ($Y_{X,\text{Lac}}$, $Y_{\text{mAb},\mu}$, $K_{\text{Asn}}$) relate to product formation and substrate affinity. These sensitivities may emerge from the time-resolved nature of the EnKF update but are less visible to a variance-based GSA. Conversely, the GSA-only parameters ($Y_{X,\text{Glc}}$, $Y_{X,\text{Gln}}$) reflect global input/output sensitivity that the EnKF does not rank as highly given its specific measurement trajectory. This agreement supports the claim that the EnKF can largely bypass the traditional GSA and reparametrisation workflow while arriving at a consistent set of influential parameters.
}

\begin{table}[ht]
\centering
\renewcommand{\arraystretch}{1.3}
\small
\setlength{\tabcolsep}{8pt}
\caption{\rev{Overlap between the ten most-converging EnKF parameters (by ensemble spread reduction) and the ten GSA-identified parameters re-estimated by Kotidis \textit{et al.}~\cite{Kotidis2019UseCultures} for Cell Line B. Seven parameters are shared (70\% overlap).}}
\label{tab:param_overlap}
\begin{tabular}{|l|l|l|}
\hline
\rowcolor{gray!30}
\textbf{GSA only (3)} & \textbf{Shared (7)} & \textbf{EnKF only (3)} \\
\hline
$Y_{X,\text{Glc}}$ & $\mu_{\text{max}}$      & $Y_{X,\text{Lac}}$ \\
$Y_{X,\text{Gln}}$ & $Lac_{\text{max1}}$     & $Y_{\text{mAb},\mu}$ \\
$m_{\text{mAb}}$    & $Lac_{\text{max2}}$     & $K_{\text{Asn}}$ \\
                     & $Y_{X,\text{Glu}}$     & \\
                     & $Y_{X,\text{Amm}}$     & \\
                     & $m_{\text{Glc}}$        & \\
                     & $Y_{X,\text{Asn}}$     & \\
\hline
\end{tabular}
\end{table}
