\label{model}

\subsection*{CHO cell growth and death}

The CHO cell growth and death kinetics were represented using Monod-type relationships, where cell proliferation depends on essential nutrients and is inhibited by metabolic by-products. The dynamic balances describing the culture volume and viable cell population are given by:\begin{align}
\frac{dV}{dt} &= F_{in} - F_{out} \\
\frac{d(VX_v)}{dt} &= (\mu - \mu_{death})V X_v - F_{out} X_v
\end{align}

\noindent where $V$ (L) denotes the culture volume, $t$ (h) the cultivation time, and $F_{in}$ and $F_{out}$ (L·h$^{-1}$) the feeding and sampling flow rates. $X_v$ (cell·L$^{-1}$) is the viable cell density, while $\mu$ and $\mu_{death}$ (h$^{-1}$) denote the specific growth and death rates.

Extracellular concentration (mM) of glucose $[Glc]$ and asparagine $[Asn]$ were modelled as the primary limiting substrates supporting cell proliferation. In the original Kotidis \textit{et al.} (2019) model, uridine $[Urd]$ acted as an inhibitory and cytotoxic factor influencing both $\mu$ and $\mu_{\text{death}}$. Since the current datasets did not involve uridine supplementation, these dependencies were excluded, and inhibitory effects were represented only through lactate $[Lac]$ and ammonia $[Amm]$ accumulation. The specific growth rate was expressed as:
\begin{align}
\mu &= \mu_{\text{max}} f_{\text{lim}} f_{\text{inh}}
\end{align}

\noindent where $\mu_{\text{max}}$ (h$^{-1}$) is the maximum specific growth rate, $f_{\text{lim}}$ represents nutrient limitation, and $f_{\text{inh}}$ represents inhibition by metabolic by-products:
\begin{align}
f_{\text{lim}} &= \frac{[Glc]}{[Glc] + K_{Glc}} \cdot \frac{[Asn]}{[Asn] + K_{Asn}} \\
f_{\text{inh}} &= \frac{KI_{Amm}}{[Amm] + KI_{Amm}} \cdot \frac{KI_{Lac}}{[Lac] + KI_{Lac}}
\end{align}

\noindent Here, $K_{Glc}$ and $K_{Asn}$ (mM) denote the Monod constants for glucose and asparagine uptake, while $KI_{Amm}$ and $KI_{Lac}$ (mM) are inhibition constants for ammonia and lactate. The cell death rate was modelled as a function of ammonia toxicity:
\begin{align}
\mu_{\text{death}} &= \mu_{\text{death,max}} \cdot \frac{[Amm]}{[Amm] + K_{d,Amm}}
\end{align}
\noindent where $\mu_{\text{death,max}}$ (h$^{-1}$) is the maximum specific death rate and $K_{d,Amm}$ (mM) the ammonia toxicity constant.


\begin{table}[h!]
\centering
\renewcommand{\arraystretch}{1.3}
\setlength{\tabcolsep}{10pt}
\small
\caption{Fitted parameters for the CHO cell growth and death model.}
\begin{tabular}{llll}
\hline
\textbf{Parameter} & \textbf{Description} & \textbf{Value} & \textbf{Unit} \\
\hline
$\mu_{max}$ & Maximum specific growth rate & $0.065$ & h$^{-1}$ \\
$\mu_{death,max}$ & Maximum specific cell death rate & $0.015$ & h$^{-1}$ \\
$K_{Glc}$ & Monod constant for glucose uptake & $14.04$ & mM \\
$K_{Asn}$ & Monod constant for asparagine uptake & $2.62$ & mM \\
$KI_{Amm}$ & Inhibition constant for ammonia & $3.17$ & mM \\
$KI_{Lac}$ & Inhibition constant for lactate & $1000$ & mM \\
$K_{d,Amm}$ & Death constant for ammonia toxicity & $14.28$ & mM \\
\hline
\end{tabular}

\end{table}


\subsection*{CHO cell metabolism}

The metabolic framework describes extracellular metabolite and amino acid dynamics through material balance equations. Each balance accounts for feeding, sampling, and cell-specific production or consumption rates as follows:

\begin{align}
\frac{d(V[Metabolite])}{dt} = F_{in}[Metabolite]_{feed} - F_{out}[Metabolite] + q_{Metabolite}V X_v
\end{align}

\noindent where $[Metabolite]$ (mM) represents the extracellular concentration, $[Metabolite]_{feed}$ (mM) the feed concentration, and $q_{Metabolite}$ (mmol·cell$^{-1}$·h$^{-1}$) the specific production or consumption rate. The following subsections summarise the kinetic expressions used for each metabolite.

\subsubsection*{Glucose}
In the original Kotidis \textit{et al.} 2019 model, glucose uptake included a galactose-dependent competitive term to capture flux diversion between the two sugars. As no galactose supplementation was present in the datasets used in this study, this term was omitted, and glucose uptake was modelled solely as a function of growth and maintenance demands:
\begin{align}
q_{Glc} = -\frac{\mu}{Y_{X/Glc}} - m_{Glc}
\end{align}
\noindent where $Y_{X/Glc}$ (cell·mmol$^{-1}$) is the yield of biomass on glucose and $m_{Glc}$ (mmol·cell$^{-1}$·h$^{-1}$) is the maintenance coefficient for glucose.

\subsubsection*{Glutamine}
The specific glutamine rate was expressed as:
\begin{align}
q_{Gln} = \frac{\mu}{Y_{X/Gln}} + q_{Amm} Y_{Gln/Amm}
\end{align}
\noindent where $Y_{X/Gln}$ (cell·mmol$^{-1}$) is the yield of biomass on glutamine and $Y_{Gln/Amm}$ (mmol·mmol$^{-1}$) is the yield of glutamine from ammonia.

\subsubsection*{Lactate}
Lactate dynamics incorporated both production and consumption behaviour:
\begin{align}
q_{Lac} =
\left( \frac{\mu}{Y_{X/Lac}} - Y_{Lac/Glc} q_{Glc} \right)
\left( \frac{Lac_{max1} - [Lac]}{Lac_{max1}} \right)
+ m_{Lac} \frac{Lac_{max2} - [Lac]}{Lac_{max2}}
\end{align}
\noindent where $Y_{X/Lac}$ (cell·mmol$^{-1}$) is the yield of biomass on lactate, $Y_{Lac/Glc}$ (mmol·mmol$^{-1}$) is the yield of lactate from glucose consumption towards the glycolysis pathway and through the synthesis of pyruvate, $Lac_{max1}$ and $Lac_{max2}$ (mM) are the activation constants for lactate consumption, and $m_{Lac}$ (mmol·cell$^{-1}$·h$^{-1}$) is the maintenance coefficient for the
participation of lactate in other metabolic pathways of the cell.

\subsubsection*{Ammonia}
Ammonia accumulation was described as:
\begin{align}
q_{Amm} = \frac{\mu}{Y_{X/Amm}} 
\end{align}
\noindent where $Y_{X/Amm}$ (cell·mmol$^{-1}$) is the yield of biomass on ammonia.

\subsubsection*{Glutamate}
Glutamate consumption was expressed as:
\begin{align}
q_{Glu} = -\frac{\mu}{Y_{X/Glu}}
\end{align}
\noindent where $Y_{X/Glu}$ (cell·mmol$^{-1}$) is the yield of biomass on glutamate.

\subsubsection*{Asparagine and Aspartate}
Asparagine uptake and its coupling with aspartate synthesis were modelled as:
\begin{align}
q_{Asn} &= -\frac{\mu}{Y_{X/Asn}} - Y_{Asn/Asp} q_{Asp} \\
q_{Asp} &= -\frac{\mu}{Y_{X/Asp}} - Y_{Asp/Asn} q_{Asn}
\end{align}
\noindent where $Y_{X/Asn}$ and $Y_{X/Asp}$ (cell·mmol$^{-1}$) are the yields of biomass on asparagine and aspartate, respectively, while $Y_{Asn/Asp}$ and $Y_{Asp/Asn}$ (mmol·mmol$^{-1}$) are the interconversion yields between asparagine and aspartate.

\subsubsection*{Monoclonal antibody synthesis}
Antibody production was described by a material balance on the extracellular antibody concentration:
\begin{align}
\frac{d(V[mAb])}{dt} &= -F_{out}[mAb] + q_{mAb}V X_v \\
q_{mAb} &= Y_{mAb,\mu}\mu + m_{mAb}
\end{align}
\noindent where $Y_{mAb,\mu}$ (mg·cell$^{-1}$) is the yield of monoclonal antibody on cell growth and $m_{mAb}$ (mg·cell$^{-1}$·h$^{-1}$) is the non-growth associated mAb production rate.

\begin{table}[h!]
\centering
\caption{Fitted parameters for the CHO cell metabolism and monoclonal antibody synthesis models.}
\renewcommand{\arraystretch}{1.3}
\setlength{\tabcolsep}{8pt}
\begin{tabular}{llll}
\hline
\textbf{Parameter} & \textbf{Description} & \textbf{Value} & \textbf{Unit} \\
\hline
$Y_{X/Glc}$ & Yield of biomass on glucose & $1.01\times10^{9}$ & cell·mmol$^{-1}$ \\
$m_{Glc}$ & Maintenance coefficient for glucose & $3.43\times10^{-11}$ & mmol·cell$^{-1}$·h$^{-1}$ \\
$Y_{X/Gln}$ & Yield of biomass on glutamine & $4.64\times10^{9}$ & cell·mmol$^{-1}$ \\
$Y_{Gln/Amm}$ & Yield of glutamine from ammonia & $0.10$ & mmol·mmol$^{-1}$ \\
$Y_{X/Lac}$ & Yield of biomass on lactate & $5.46\times10^{7}$ & cell·mmol$^{-1}$ \\
$Y_{Lac/Glc}$ & Yield of lactate from glucose & $1.56$ & mmol·mmol$^{-1}$ \\
$Lac_{max1}$ & Activation constant for lactate consumption & $21.20$ & mM \\
$Lac_{max2}$ & Activation constant for lactate consumption & $16.00$ & mM \\
$m_{Lac}$ & Maintenance coefficient for lactate & $1.87\times10^{-10}$ & mmol·cell$^{-1}$·h$^{-1}$ \\
$Y_{X/Amm}$ & Yield of biomass on ammonia & $2.36\times10^{9}$ & cell·mmol$^{-1}$ \\
$Y_{X/Glu}$ & Yield of biomass on glutamate & $1.46\times10^{10}$ & cell·mmol$^{-1}$ \\
$Y_{X/Asn}$ & Yield of biomass on asparagine & $7.68\times10^{8}$ & cell·mmol$^{-1}$ \\
$Y_{Asn/Asp}$ & Yield of asparagine from aspartate & $0.10$ & mmol·mmol$^{-1}$ \\
$Y_{X/Asp}$ & Yield of biomass on aspartate & $3.59\times10^{9}$ & cell·mmol$^{-1}$ \\
$Y_{Asp/Asn}$ & Yield of aspartate from asparagine & $0.13$ & mmol·mmol$^{-1}$ \\
$Y_{mAb,\mu}$ & Yield of monoclonal antibody on cell growth & $3.39$ & pg·cell$^{-1}$ \\
$m_{mAb}$ & Non-growth associated mAb production rate & $4.10\times10^{-1}$ & pg·cell$^{-1}$·h$^{-1}$ \\
\hline
\end{tabular}
\end{table}



